First Order Alternation

### 2018-09-21 quantifiers/draft.tex
ok

The problem of checking emptiness of a given automata is undecidable,
even for automata with predicates of arity two, whose transition rules
use only equalities and disequalities, with no transition quantifiers
\cite{Farzan15}. Since even such simple classes of alternating
automata have no general decision procedure for emptiness, we use an
abstraction-refinement semi-algorithm based on \emph{lazy annotation}
\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
procedure systematically explores the set of execution traces of a
program in search for a counterexample. When a spurious counterexample
is found, the search backtracks and the program path is annotated with
a set of learned facts, that excludes the path from future
searches. These facts are interpolants extracted from the proof of
spuriousness of the given (finite) path. The emptiness checking
semi-algorithm uses moreover a coverage relation to ensure that a
spurious counterexample is never explore twice, thus making progress
towards either finding a real counterexample or a proof of emptiness.

Our implementation is a version of the \impact~ lazy annotation
semi-algorithm \cite{McMillan06}. The detailed pseudocode has been
given for a simpler model of boolean alternating automata extended
with variables, in \cite{IosifXu18} and is reported, for
self-completeness, in Appendix \ref{app:pseudocode}. For the time
being, we shall content ourselves of a high-level definition of the
data structure used to prove emptiness of an automaton.

@@

\begin{definition}\label{def:unfolding}
  An \emph{unfolding} of an automaton $\A =
  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is a finite partial mapping $U :
  \Sigma^* \rightharpoonup_{\mathit{fin}} \posforms(Q,X)$, such
  that: \begin{compactenum}
  \item\label{it1:unfolding} $\dom(U)$ is a finite prefix-closed
    complete set,
    %
  \item\label{it2:unfolding} $U(\epsilon) = \iota$, and
    %
  \item\label{it3:unfolding} for each sequence $\alpha a \in \dom(U)$,
    such that $\alpha \in \Sigma^*$ and $a \in \Sigma$:
    \[\stamp{U(\alpha)}{0} \wedge
    \bigwedge_{q(\vec{y}) \arrow{a(X)}{} \psi} \forall y_1 \ldots
    \forall y_{\#q} ~.~ \stamp{q}{0}(\vec{y}) \arrow{a(X)}{}
    \stamp{\psi}{1} \models \stamp{U(\alpha a)}{1}\]
  \end{compactenum}
  Moreover, $U$ is \emph{safe} if for each $\alpha \in \dom(U)$, the
  formula $U(\alpha) \wedge \bigwedge_{q \in Q \setminus F} \forall
  y_1 \ldots \forall_{y_{\#(q)}} ~.~ q(\vec{y}) \rightarrow \bot$ is
  unsatisfiable. 
\end{definition}

Lazy annotation semi-algorithms \cite{McMillan06,McMillan14}
essentially build unfoldings of automata trying to discover
counterexamples for emptiness. If the automaton $\A$ in question is
non-empty, a systematic enumeration of the input event
sequences\footnote{For instance, using breadth-first search.} from
$\Sigma^*$ will suffice to discover a word $w \in \lang{\A}$, provided
that the first-order theory of the data domain $\Data$ is decidable
(Lemma \ref{lemma:acceptance}). However, if $\lang{\A} = \emptyset$,
the enumeration of input event sequences may, in principle, run
forever. The typical way of fighting this divergence problem is to
define a \emph{coverage} relation between the nodes of the unfolding
tree.

@@

A lazy annotation semi-algorithm will stop and report emptiness
provided that it succeeds in building a closed and safe unfolding of
the automaton. For instance, the \impact~ semi-algorithm
\cite{McMillan06} upon which our procedure is based, continuosly
expands the domain of the unfolding, while partially recomputing its
labeling and the coverage, and stops when a safe complete unfolding
has been found. The following theorem proves the soundness of this
method:

\begin{theorem}\label{thm:soundness}
  Given an automaton $\A$, if $\A$ has a nonempty safe closed
  unfolding then $\lang{A} = \emptyset$. 
\end{theorem}
\proof{

@@

As mentioned above, we check emptiness of first-order alternating
automata using the same method previously used to check emptiness of a
simpler model of alternating automata, which uses boolean constants
for control states and whose transition rules have no quantifiers
\cite{IosifXu18}. The higher complexity of the automata model
considered here, manifests itself within the interpolant generation
procedure, used to label the unfoldings. We shall discuss generation
of interpolants in the remainder of this section.

\subsection{Interpolant Generation}\label{sec:interpolants}

>>>>

The problem of checking emptiness of a given automata is undecidable,
even for automata with predicates of arity two, whose transition rules
use only equalities and disequalities, with no transition quantifiers
\cite{Farzan15}. Since even such simple classes of alternating
automata have no general decision procedure for emptiness, we use an
abstraction-refinement semi-algorithm based on \emph{lazy annotation}
\cite{McMillan06,McMillan14}.

In a nutshell, a lazy annotation procedure systematically explores the
set of execution paths (in our case, sequences of input events) in
search of an accepting execution. Each path has a corresponding path
formula that defines all words accepted along that path. If the
current path formula is satisfiable, the automaton accepts a word and
is not empty. Otherwise, the path is \emph{spurious}. When a spurious
path is encountered, the search backtracks and the path is annotated
with a set of learned facts, that marks this path as spurious in all
future searches. The semi-algorithm uses moreover a coverage relation
between paths, that optimize the search (the continuations of already
covered paths are never explored) and sometimes provide a sound
termination argument, when the automaton is empty.

We check emptiness of first-order alternating automata using a version
of the \impact~ lazy annotation semi-algorithm \cite{McMillan06}. The
detailed pseudocode has been given for a simpler model of boolean
alternating automata extended with variables \cite{IosifXu18}. To save
space, we do not present the details of this algorithm (given in
Appendix \ref{app:pseudocode}) and shall content ourselves of several
high-level definitions.

@@

\begin{definition}\label{def:unfolding}
  An \emph{unfolding} of an automaton $\A =
  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is a finite partial mapping $U :
  \Sigma^* \rightharpoonup_{\mathit{fin}} \posforms(Q,\emptyset)$,
  such that: \begin{compactenum}
  \item\label{it1:unfolding} $\dom(U)$ is a finite prefix-closed
    complete set,
    %
  \item\label{it2:unfolding} $U(\epsilon) = \iota$, and
    %
  \item\label{it3:unfolding} for each sequence $\alpha a \in \dom(U)$,
    such that $\alpha \in \Sigma^*$ and $a \in \Sigma$:
    \[\stamp{U(\alpha)}{0} \wedge
    \bigwedge_{q(\vec{y}) \arrow{a(X)}{} \psi} \forall y_1 \ldots
    \forall y_{\#q} ~.~ \stamp{q}{0}(\vec{y}) \rightarrow
    \stamp{\psi}{1} \models \stamp{U(\alpha a)}{1}\]
  \end{compactenum}
  Moreover, $U$ is \emph{safe} if for each $\alpha \in \dom(U)$, the
  formula $U(\alpha) \wedge \bigwedge_{q \in Q \setminus F} \forall
  y_1 \ldots \forall_{y_{\#(q)}} ~.~ q(\vec{y}) \rightarrow \bot$ is
  unsatisfiable. 
\end{definition}

Lazy annotation semi-algorithms \cite{McMillan06,McMillan14} build
unfoldings of automata trying to discover counterexamples for
emptiness. If the automaton $\A$ in question is non-empty, a
systematic enumeration of the input event sequences\footnote{For
  instance, using breadth-first search.} from $\Sigma^*$ will suffice
to discover a word $w \in \lang{\A}$, provided that the first-order
theory of the data domain $\Data$ is decidable (Lemma
\ref{lemma:acceptance}). However, if $\lang{\A} = \emptyset$, the
enumeration of input event sequences may, in principle, run
forever. The typical way of fighting this divergence problem is to
define a \emph{coverage} relation between the nodes of the unfolding
tree.

@@

A lazy annotation semi-algorithm will stop and report emptiness
provided that it succeeds in building a closed and safe unfolding of
the automaton. Notice that, by Definition \ref{def:coverage}, for any
three nodes of an unfolding $U$, say $\alpha,\beta,\gamma \in
\dom(U)$, if $\alpha \prec \beta$ and $\alpha \cover \gamma$, then
$\beta \cover \gamma$ as well. As we show next (Theorem
\ref{thm:soundness}), there is no need to expand covered nodes,
because, intuitively, there is a word $w \in \lang{\A}$ such that
$\alpha \preceq \event{w}$ and $\alpha \cover \gamma$ only if there
exists another word $u \in \lang{\A}$ such that $\gamma \preceq
\event{u}$.

An unfolding node $\alpha \in \dom(U)$ is said to be \emph{spurious}
if and only if $\accform{\alpha}$ is unsatisfiable. In this case, we
change (refine) the labels $U(\beta)$ of (some of the) prefixes $\beta
\prefix \alpha$, including $\alpha$ itself, such that $U(\alpha)$
becomes $\bot$, thus indicating that there is no real execution of the
automaton along that input event sequence. As a result of the change
of labels, if a node $\gamma \prefix \alpha$ used to cover another
node from $\dom(U)$, it might not cover it with the new
label. Therefore, the coverage relation has to be recomputed after
each refinement of the labeling. For a detailed presentation of the
emptiness procedure, we refer the reader to \cite{IosifXu18} or
Appendix \ref{app:pseudocode}.

The semi-algorithm stops when (and if) a safe complete unfolding has
been found. The following theorem proves the soundness of this
termination argument:

\begin{theorem}\label{thm:soundness}
  Given an automaton $\A$, if $\A$ has a nonempty safe closed
  unfolding then $\lang{A} = \emptyset$. 
\end{theorem}
\proof{ Let $U$ be a safe and complete unfolding of $\A$, such that
  $\dom(U) \neq \emptyset$. Suppose, by contradiction, that there
  exists a word $w \in \lang{A}$ and let $\alpha \isdef
  \event{w}$.

  Since $w \in \lang{A}$, by Lemma \ref{lemma:acceptance}, that there
  exists an interpretation $\I$ such that $\I,\data{w} \models
  \accform{\alpha}$. If $\alpha \in \dom(U)$ then one can show, by
  induction on the length $n\geq0$ of $w$, that $\pathform{\alpha}
  \models \stamp{U(\alpha)}{n}$, thus $\I,\data{w} \models
  \stamp{U(\alpha)}{n}$. But $\I,\data{w} \models \bigwedge_{q \in Q
    \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n}(\vec{y}) \rightarrow \bot$, hence $\stamp{U(\alpha)}{n}
  \wedge \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
  y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$. By renaming
  $\stamp{q}{n}$ with $q$ in the previous formula, we obtain that
  $U(\alpha) \wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~
  q(\vec{y}) \rightarrow \bot$ is satisfiable, thus $U$ is not safe,
  contradiction.

@@

As mentioned above, we check emptiness of first-order alternating
automata using the same method previously used to check emptiness of a
simpler model of alternating automata, which uses boolean constants
for control states and whose transition rules have no quantifiers
\cite{IosifXu18}. The higher complexity of the automata model
considered here, manifests itself within the interpolant generation
procedure, used to refine the labeling of the unfolding. We discuss
generation of interpolants in the next section.

\section{Interpolant Generation}\label{sec:interpolants}
